Measuring Directionality in Double-Beta Decay and Neutrino Interactions with Kiloton-Scale Scintillation Detectors

### 2014-04-21 Direction_v3_withLindley.tex
Add old versions of paper. --Lindley

@@

The detector geometry is a sphere of 6.5~m diameter filled with scintillator. FIG.~\ref{detector_view} shows the geometry and the Cherenkov light from an example $^{116}$Cd $0\nu\beta\beta$ event. The default
scintillator properties have been chosen to match the KamLAND
scintillator \cite{kamland2003}: 80~\% n-Dodecane, 20~\% Pseudocumene
(1,2,4-Trimethylbenzene) and 1.52~g/l PPO (2,5-Diphenyloxazole). The
scintillator properties implemented in the simulation include the
atomic composition and density ($\rho$ = 0.78~g/ml), the
wavelength-dependent attenuation length \cite{tajimaMaster} and
refractive index \cite{OlegThesis}, the scintillation emission
spectrum\cite{tajimaMaster}, emission rise time ($\tau_r$ = 1.0~ns) and emission decay
time constants ($\tau_{d1}$ = 6.9~ns and $\tau_{d2}$ = 8.8~ns with
relative weights of 0.87 and 0.13)
\cite{tajimaThesis}, scintillator light yield (LY, 9030 photons/MeV)
and the Birks constant ($kB$ $\approx$ 0.1~mm/MeV)
\cite{ChrisThesis}. Variations from the baseline KamLAND case are
discussed below. Re-emission of absorbed photons in the scintillator
bulk volume and scattering have not yet been included, but are not
expected to change the conclusions here.

@@

\bibliography{DirectionBibliography}

>>>>

In $0\nu\beta\beta$, the electrons emerge with a combined energy equal to the endpoint of the particular isotope's Q-value. The individual electrons come out with a distribution of energies and angular correlations. The most probable being equal division of energy between the back-to-back electrons\cite{phasespace} as is shown in FIG.~\ref{detector_view} for an example $^{116}$Cd $0\nu\beta\beta$ event. Since the decay half-life is proportional to the phase-space factor, isotope's with higher Q-values are preferred and due to backgrounds from the daughters of the $^{238}$U and $^{232}$Th decay chains those with Q-values at or above 2.6~MeV at most often considered for $0\nu\beta\beta$ searches. There are hundreds of candidate $0\nu\beta\beta$ isotopes\cite{tabledbb}, but only a handful with large Q-values.  Most of the high Q-value candidates have been considered as a dopant for a liquid scintillator: $^{150}$Nd(Q=3.367~MeV)\cite{minfang,nd1}, $^{96}$Zr(Q=3.350~MeV)\cite{zr1},  $^{100}$Mo(Q=3.034~MeV)\cite{mo1}, $^{82}$Se(Q=3.350~MeV)\cite{qdot}, $^{116}$Cd(Q=2.802~MeV)\cite{qdot, cd1}, $^{130}$Te(Q=2.533~MeV)\cite{qdot, biller}, $^{136}$Xe(Q=2.479~MeV)\cite{KZ0nu} and $^{124}$Sn(Q=2.29~MeV)\cite{sn1}. Xenon gas readily dissolves into liquid scintillator. For the other isotopes, the correct organometallic compound needs to be found that produces a stable scintillator with a long attenuation length in the wavelength region of interest. Recently, nanocrystals called quantum dots have been explored as an alternative method for doping scintillator\cite{qdot}. Quantum dots have interesting optical properties which are discussed further in Section~\ref{scintillator_emission_sec}.

The reconstruction of the direction of the two simultaneous emitted $\sim$1.5~MeV electrons from the $0\nu\beta\beta$ decay is quite challenging. In this paper, we decided to start our directional studies with 5~MeV single electrons since reconstruction of direction in this energy range had never been achieved. This is a logical energy to use since this is where historically scintillating detectors become more practical than water Cherenkov detectors.  This is also an energy that is relevant for studies of $^{8}$B solar neutrinos and neutrino-electron elastic scattering measurement.

@@

The detector geometry is a sphere of 6.5~m diameter filled with scintillator, see FIG.~\ref{detector_view} shows the geometry and the Cherenkov light from an example $^{116}$Cd $0\nu\beta\beta$ event. The default
scintillator properties have been chosen to match the KamLAND
scintillator \cite{kamland2003}: 80~\% n-Dodecane, 20~\% Pseudocumene
(1,2,4-Trimethylbenzene) and 1.52~g/l PPO (2,5-Diphenyloxazole). The
scintillator properties implemented in the simulation include the
atomic composition and density ($\rho$ = 0.78~g/ml), the
wavelength-dependent attenuation length \cite{tajimaMaster} and
refractive index \cite{OlegThesis}, the scintillation emission
spectrum\cite{tajimaMaster}, emission rise time ($\tau_r$ = 1.0~ns) and emission decay
time constants ($\tau_{d1}$ = 6.9~ns and $\tau_{d2}$ = 8.8~ns with
relative weights of 0.87 and 0.13)
\cite{tajimaThesis}, scintillator light yield (LY, 9030 photons/MeV)
and the Birks constant ($kB$ $\approx$ 0.1~mm/MeV)
\cite{ChrisThesis}. Variations from the baseline KamLAND case are
discussed below. Re-emission of absorbed photons in the scintillator
bulk volume and scattering have not yet been included, but are not
expected to change the conclusions here.

@@

\bibliography{DirectionBibliography_withLindley}

### 2014-04-21 Direction_v7_submit.tex
Add old versions of paper. --Lindley

\begin{figure*}[tbh]
\begin{center}
        \subfigure[ ~Default
        simulation.]{\includegraphics[scale=0.295]{graphs/6p5Meter_5MeVElectrons_Bialkali_KamlandScintSpec_TIME.pdf}}
        \subfigure[ ~Increased TTS
        (1.28~ns).]{\includegraphics[scale=0.295]{graphs/6p5Meter_5MeVElectrons_Bialkali_KamlandScintSpec_1p28nsTTS_TIME.pdf}}
        \subfigure[ ~Red-sensitive
        photocathode.]{\includegraphics[scale=0.295]{graphs/6p5Meter_5MeVElectrons_RedSensitiveQE_KamlandScintSpec_TIME.pdf}}
        \caption[]{Photoelectron (PE) arrival times after application
        of the transit-time spread (TTS) for the simulation
        of 1000 electrons (5~MeV) with different values of the TTS and
        wavelength response. PEs from
        Cherenkov light (black, solid line) and scintillation light
        (red, dotted line) are
        compared. The dash-dotted vertical line illustrates a time cut at
        34.0~ns. (a) Default simulation: bialkali photocathode and TTS =
        0.1~ns ($\sigma$). After the 34.0~ns time cut we get 171~PEs
        from scintillation and 108~PEs from Cherenkov light. (b)
        Default simulation settings except for TTS = 1.28~ns (KamLAND
        17 in. PMTs). After the 34.0~ns time cut we get 349~PEs from
        scintillation and 88~PEs from Cherenkov light. (c) Default
        simulation settings except for a GaAsP photocathode. After the
        34.0~ns time cut we get 226~PEs from scintillation and 229~PEs
        from Cherenkov light. \label{time_plots_comparison}}
        \end{center}
\end{figure*}

@@

\begin{figure}
        \begin{center}
        \includegraphics[scale=0.40]{graphs/cos_psi_34_h.pdf}
        \caption[]{The angular distribution of photoelectron hits
        relative to the original electron direction, $\cos(\Psi) =
        x_{hit}/|\vec{r}_{hit}|$. The sample consists of 1000 events
        with a 5~MeV electron produced at the detector center. Default
        simulation settings are used and both Cherenkov and
        scintillation light are included. The $t\leq$ 34.0~ns cut is applied.} 
        \label{Cherenkov_cone}
        \end{center}
\end{figure}

>>>>

\begin{figure*}[tbh]
\begin{center}
        \subfigure[ ~Default
        simulation.]{\includegraphics[scale=0.295]{6p5Meter_5MeVElectrons_Bialkali_KamlandScintSpec_TIME.pdf}}
        \subfigure[ ~Increased TTS
        (1.28~ns).]{\includegraphics[scale=0.295]{6p5Meter_5MeVElectrons_Bialkali_KamlandScintSpec_1p28nsTTS_TIME.pdf}}
        \subfigure[ ~Red-sensitive
        photocathode.]{\includegraphics[scale=0.295]{6p5Meter_5MeVElectrons_RedSensitiveQE_KamlandScintSpec_TIME.pdf}}
        \caption[]{Photoelectron (PE) arrival times after application
        of the transit-time spread (TTS) for the simulation
        of 1000 electrons (5~MeV) with different values of the TTS and
        wavelength response. PEs from
        Cherenkov light (black, solid line) and scintillation light
        (red, dotted line) are
        compared. The dash-dotted vertical line illustrates a time cut at
        34.0~ns. (a) Default simulation: bialkali photocathode and TTS =
        0.1~ns ($\sigma$). After the 34.0~ns time cut we get 171~PEs
        from scintillation and 108~PEs from Cherenkov light. (b)
        Default simulation settings except for TTS = 1.28~ns (KamLAND
        17 in. PMTs). After the 34.0~ns time cut we get 349~PEs from
        scintillation and 88~PEs from Cherenkov light. (c) Default
        simulation settings except for a GaAsP photocathode. After the
        34.0~ns time cut we get 226~PEs from scintillation and 229~PEs
        from Cherenkov light. \label{time_plots_comparison}}
        \end{center}
\end{figure*}

@@

\begin{figure}
        \begin{center}
        \includegraphics[scale=0.40]{cos_psi_34_h.pdf}
        \caption[]{The angular distribution of photoelectron hits
        relative to the original electron direction, $\cos(\Psi) =
        x_{hit}/|\vec{r}_{hit}|$. The sample consists of 1000 events
        with a 5~MeV electron produced at the detector center. Default
        simulation settings are used and both Cherenkov and
        scintillation light are included. The $t\leq$ 34.0~ns cut is applied.} 
        \label{Cherenkov_cone}
        \end{center}
\end{figure}
